Do larger catchments respond different to forest cover change? Re-analysing a global data set.

### 2021-11-23 paper/Forest_and_Water/Forest_and_Water.Rmd
further paper updates and data correction

---
title: Do larger catchments respond different to forest cover change? Re-analysing a global data set.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: The University of Sydney, INIA
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
    footnote: 2
address:
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: INIA, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Departamento de Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
abstract: |
  This is the abstract.
  It consists of two paragraphs.
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
editor_options: 
  chunk_output_type: console
---

@@

Several additional data points from watershed studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011)

@@

@@

In this model, no direct interactions are assumed. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and non-hysteretic. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data is converted to positive values, and an additional column (*sign_{forest cover}*) is added that indicates whether it was a forest cover increase or decrease. 
A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Q \%$. This means that a smooth function $s()$ is applied to the $Z_i$ variables.

@@

The overall dataset contains `r nrow(Zhang_all)` observations of changes in flow. The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction.

```{r, echo = F}
Zhang_all %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() + facet_wrap(~variable, scales = "free")
```

```{r, echo = F}
Zhang_all2 <- Zhang_all %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "decrease", "increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))

# quick plot to show differences
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + scale_colour_manual(name = "direction of forest change", values = c("increase" = "blue", "decrease" = "red")) +
  xlab("Change in forest cover %") + ylab("Change in flow %")

@@

This immediately shows that for the data related to forest decreases, there is almost always a positive flow change. In other words, flow almost always increased. However, for increases in forest cover, this is not the case, and flow can both increase and decrease. However in both cases the variability in the reported change in flow increases with the increase in forest cover change.

@@

```{r model1, echo = F}
linear_model_simple <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
                             Forest_Sign, 
                           data = Zhang_all2)
summary(linear_model_simple)
gam.check(linear_model_simple)

@@

```{r, echo =F}
pander(summary(linear_model_simple$p.table))
```

@@

linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm, data = Zhang_all2)
summary(linear_model2)
gam.check(linear_model2)
```

```{r, echo = F}
pander(summary(linear_model2)$p.table)
```

@@

```{r model3, echo = F}
model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
summary(model3_includeLatLong)
gam.check(model3_includeLatLong)
```

```{r, echo =F}
pander(summary(model3_includeLatLong)$p.table)
```

@@

The previous analysis suggests an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017. Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics). This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variables.

@@

```{r model4, echo = F}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)
model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 +  
                        Latitude + Longitude + Dryness, data = Zhang_all2)
summary(model4_dryness)
gam.check(model4_dryness)
```

```{r, echo =F}
pander(summary(model4_dryness)$p.table)
```

@@

<!-- model 2b also include categorical variables -->
```{r model5, echo = F}
Zhang_all2 <- Zhang_all2 %>%
  mutate(length = To - From)

model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2) + s(Dryness) + length +  Assessment_technique + Latitude ,  data = Zhang_all2 %>% filter(Dryness < 4))
summary(model5_all)
gam.check(model5_all)
plot(model5_all)
```
<!-- model 2c include smooth variables -->

```{r model2c, echo = F}

model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + te(Dryness,Latitude, k =5) + Precip_data_type + Assessment_technique + Forest_type + Hydrological_regime, data = Zhang_all2)
summary(model6_all)
gam.check(model6_all)
plot(model6_all)
```

```{r}
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Dryness)) + geom_point()
```

```{r}
Zhang_all2 %>%
  ggplot(aes(Assessment_technique,Dryness)) + geom_boxplot()
```

```{r}
Zhang_all2 %>%
  filter(Dryness > 4)

@@

```{r}
Zhang_all %>%
  ggplot(aes(Assessment_technique,log10(Area_km2))) + geom_boxplot()
```

```{r}
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), shape = Forest_Sign)) + geom_point()
```
```{r}
tiff("flow_forest_byArea.tiff", width = 2500, height = 1800, res = 300)
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
             shape = Forest_Sign)) + geom_point(alpha = 0.5) +
  theme_bw() + ylab("% change in flow") +
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size = rel(1.5))) +
  xlab("% change in forestry") + #scale_y_log10() +
  scale_size_continuous(name = "log10(Area in km2)") +
  scale_colour_discrete(name = "Assessment Technique") +
  scale_shape_discrete(name = "Forest cover direction")
dev.off()
```

```{r}
Zhang_all2 %>%
  ggplot(aes(Longitude, Latitude, colour = DeltaF_perc, size = DeltaQf_perc/100 )) + geom_point(alpha = 0.5)
```

@@

```{r}
Zhang_all %>%
  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
  scale_x_log10()
total <- nrow(Zhang_all)
length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
```
```{r}
Zhang_all2 %>%
  ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
```
Essentially, the analysis shows at the moment that their is no evidence that the size of the catchment influences the

>>>>

---
title: Do larger catchments respond different to forest cover change? Re-analysing a global data set.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: The University of Sydney, INIA
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA
    footnote: 2
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
    footnote: 2
address:
  - code: The University of Sydney
    address: School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: INIA, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Departamento de Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
abstract: |
  This is the abstract.
  It consists of two paragraphs.
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
editor_options: 
  chunk_output_type: console
---

@@

Several additional data points from watershed studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to Watershed #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired watershed studies in French Guyana, and in which the actual change in forest cover is not recorded.

@@

<!--remove the Amazon from the analysis -->
```{r}
Zhang_all <- Zhang_all %>%
  filter(`Watershed #` > 1)
```

@@

Here $X_i$ are factorial variables, while $Z_i$ are continous variables. The model assumes no direct interactions and all variables are additive. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and non-hysteretic. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data is converted to positive values, and an additional column (*sign_{forest cover}*) is added that indicates whether it was a forest cover increase or decrease. 
A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have a linear or non-linear relationship with $\Delta Q \%$. This means that a smooth function $s()$ is applied to the $Z_i$ variables.

@@

The overall dataset contains `r nrow(Zhang_all)` observations of changes in flow. The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and Area. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Smaller watersheds dominate the database with 42% of the data from watersheds < 1 km^2^ and 65% of the data for watersheds < 10 km^2^.

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
Zhang_all %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, Area_km2, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("Area_km2" = "Area (km2)",
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Flow (%)",
      "E0" = "Annual Potential E (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"))) +   
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

```{r data_graphs, echo =F, fig.cap = "Overview of the distributions of some of the variables in the data set", out.width= "90%"}
include_graphics("DataExploration.png")
```

```{r, echo = F}
Zhang_all2 <- Zhang_all %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "decrease", "increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))
```

```{r, echo = F, warning = F, message=F, eval =F}
# plot to show the difference
png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + scale_colour_manual(name = "direction of forest change", values = c("increase" = "blue", "decrease" = "red")) +
  xlab("Change in forest cover %") + ylab("Change in flow %")
dev.off()

@@

```{r increase_decrease, echo =F, fig.cap = "Changes in flow as a function of increases and decreases in forest cover", out.width= "90%"}
include_graphics("Increase_decrease.png")
```

This shows that for the data related to forest decreases, there is almost always a positive flow change. In other words, flow almost always increased. However, for increases in forest cover, this is not the case, and flow can both increase and decrease. However in both cases the variability in the reported change in flow increases with the increase in forest cover change.

@@

```{r model1, echo = F}
linear_model_simple <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
                             Forest_Sign, 
                           data = Zhang_all2)
#summary(linear_model_simple)
#gam.check(linear_model_simple)

@@

```{r, echo =F}
pander(round(summary(linear_model_simple)$p.table,2))
```

@@

linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm, data = Zhang_all2)
#summary(linear_model2)
#gam.check(linear_model2)
```

```{r, echo = F}
pander(round(summary(linear_model2)$p.table,2))
```

@@

```{r model3, echo = F}
model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm + Latitude + Longitude, data = Zhang_all2)
#summary(model3_includeLatLong)
#gam.check(model3_includeLatLong)
```

```{r, echo =F}
pander(round(summary(model3_includeLatLong)$p.table,2))
```

@@

The previous analysis suggests an influence of radiation on evapotranspiration, and most likely related to the dryness index, as also indicated in @zhang2017. Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics). This model introduces the dryness index as a linear variable and drops the annual average precipitation as a variable.

@@

```{r model4, echo = F}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)
model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 +  
                        Latitude + Longitude + Dryness, data = Zhang_all2)
#summary(model4_dryness)
#gam.check(model4_dryness)
```

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2))
```

@@

However, the result also indicates possible issues with the data, some of the Dryness values are very large (> 4) and these values have high leverage in the data. These catchment are listed in Table XX:

```{r, echo = F}
pander(Zhang_all2 %>% filter(Dryness > 4) %>% select(Latitude, Longitude,`Watershed name`))
```

```{r model5, echo = F}
Zhang_all2 <- Zhang_all2 %>%
  mutate(length = To - From)

model5_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + s(Area_km2) + s(Dryness) + length +  Assessment_technique + Latitude ,  data = Zhang_all2 %>% filter(Dryness < 4))
#summary(model5_all)
#gam.check(model5_all)
#plot(model5_all)
```
<!-- model 2c include smooth variables -->

```{r model6_all, echo = F}

model6_all <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts" ) + 
                    Precip_data_type + Assessment_technique + Forest_type +
                    Hydrological_regime, data = Zhang_all2)
summary(model6_all)
#gam.check(model6_all)
#plot(model6_all)
```

```{r model6_reduc}
model6_reduc <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + 
                    s(Dryness, bs="ts" ) + s(Latitude, bs="ts" ) + 
                    Assessment_technique +
                    Hydrological_regime, data = Zhang_all2)
summary(model6_reduc)

```{r, echo = F}
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Dryness)) + geom_point()
```

```{r, echo = F}
Zhang_all2 %>%
  ggplot(aes(Assessment_technique,Area_km2)) + geom_boxplot()
```

```{r, eval =F}
Zhang_all2 %>%
  filter(Area_km2 > 3e+06)

@@

```{r, eval =F}
Zhang_all %>%
  ggplot(aes(Assessment_technique,log10(Area_km2))) + geom_boxplot()
```

```{r, echo = F}
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), shape = Forest_Sign)) + geom_point()
```
```{r, echo = F}
png("flow_forest_byArea.png", width = 2500, height = 1800, res = 300)
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
             shape = Forest_Sign)) + geom_point(alpha = 0.5) +
  theme_bw() + ylab("% change in flow") +
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size = rel(1.5))) +
  xlab("% change in forestry") + #scale_y_log10() +
  scale_size_continuous(name = "log10(Area in km2)") +
  scale_colour_discrete(name = "Assessment Technique") +
  scale_shape_discrete(name = "Forest cover direction")
dev.off()
```

```{r, eval =F}
Zhang_all2 %>%
  ggplot(aes(Longitude, Latitude, colour = DeltaF_perc, size = DeltaQf_perc/100 )) + geom_point(alpha = 0.5)
```

@@

```{r, eval =F}
Zhang_all %>%
  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
  scale_x_log10()
total <- nrow(Zhang_all)
length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
```
```{r, eval =F}
Zhang_all2 %>%
  ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
```
Essentially, the analysis shows at the moment that their is no evidence that the size of the catchment influences the
